Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-04-20 manuscript/manuscript.Rmd
updated some references.
added some references.

Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008]. The controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter. In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]

@@

Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal. This approach is a power and elegant yet limited by the curse of expertise [@hinds1999]. Like chaine operatoire, it cannot handles variation very well. To some extent it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps. Novices has a different sets of perception on the causal structure of how certain behaviors will modify the raw materials, leading to over-imitation. Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014].

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel. However, it has never been formally recognized as a legitimate research method in experimental archaeology.

>>>>

Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. The controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter. In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]

@@

Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal. This approach is a power and elegant yet limited by the curse of expertise [@hinds1999]. Like chaine operatoire, it cannot handles variation very well. To some extent it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps. Novices has a different sets of perception on the causal structure of how certain behaviors will modify the raw materials, leading to over-imitation. Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014], or rotation analysis [@muller2023]

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel. However, it has never been formally recognized as a legitimate research method in experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.
